---
title: The Mandelbrot set in R
author: Daniel Burrell
date: '2020-03-18'
slug: the-mandelbrot-set-in-r
categories:
  - R
  - fractals
  - dynamics
tags:
  - fractals in R
description: ''
topics: []
---

## Introduction

This being my first post to Dandelbrot.com, I thought it only fitting to post something about the Mandelbrot set. I determined to code up a mandelbrot set fractal generator in R. After a little digging I came across [this R-bloggers post](https://www.r-bloggers.com/the-mandelbrot-set-in-r/) from 2014. I figured, rather than reinvent the wheel, I might as well just borrow code from here, with all due kutos to Myles Harrison, the original author. 

The Mandelbrot set and its colourful depictions as the Mandelbrot fractal image is probably one of the better known fractal objects. It begins with complex numbers of the form $z = a + bi$, where $a$ and $b$ are real numbers, and $i$ is the imaginary unit satisfying $i^2 = -1$. The Mandelbrot set is generated as follows:

1. Initialise a complex set of numbers in the complex plane that are all $z=0$. 
2. Iterate the formula: $z_{n+1}=z_n^2 + c$ such that $c$ is a set of complex numbers that fills the complex plane. 
3. The Mandelbrot set is the set where $z$ remains bounded for all $n$.

## Implementation in R
The idea here is to pick some parameters about the space (such as range and resolution for $x$ and $y$, number of iterations, etc.) and then iterate. [Myles Harrison](https://everydayanalytics.ca/2014/12/the-mandelbrot-set-in-r.html) builds up the code by taking the reader through an iterative process, starting from a naive code using loops, to a vectorised version, through to a functional version. I'm not going to go through the process here, but I will recode the final vectorised function below:
```{r}
mandelbrot_generator <- function(
    p = 2, 
    q = 1,
    xmin = -2.1, # minimum x value
    xmax = 0.8,  # maximum x value
    nx = 500, 
    ymin = -1.3, # minimum y value
    ymax = 1.3,  # maximum y value
    ny = 500,
    n = 100, 
    showplot = TRUE, # If TRUE then display image,
    showvals = FALSE,
    cols = colorRampPalette(c("black","cyan","cyan3","black"))(11)) 
{
    
    # variables
    x <- seq(xmin, xmax, length.out=nx)
    y <- seq(ymin, ymax, length.out=ny)
    c <- outer(x,y*1i,FUN="+")
    z <- matrix(0.0, nrow=length(x), ncol=length(y))
    k <- matrix(0.0, nrow=length(x), ncol=length(y))
    
    for (rep in 1:n) { 
        index <- which(Mod(z) < 2)
        z[index] <- z[index]^p + c[index]*q
        k[index] <- k[index] + 1
    }
    
    if (showplot==TRUE) { image(x,y,k,col=cols, xlab="Re(c)", ylab="Im(c)")}
    if (showvals==TRUE) {return(k)}
    
}
```
Running this function with all the arguments as set (note that these differ somewhat to the arguments in the original) leads to the following:
```{r}
mandelbrot_generator(p=2, q=1)
```
This is the code I used to create the logo for Dandelbrot.com. I find fractal geometry hugely pleasing. I love dynamical systems and the idea that such simple iterative functions can produce such complex and intricate behaviour. The Mandelbrot set is just a commonly known embodiment of this. I urge you to play around with the code and see what else you can come up with. For example, what adjustments to the basic iterative equation can you make, and how does it impact the result? Below, I have tweaked the original `mandelbrot_generator()` function to enable general powers:
\[
z_{n+1} = z_n^p + c^q.
\]
When $p=4$ and $q=0.5$ and the necessary adjustments to the plotting window are made, the following image is generated:
```{r}
mandelbrot_generator(p=4, q=0.5, xmin = -2.8, xmax = 2, ymin = -2.5, ymax = 2.5, nx=1000, ny=1000, n=75)
```

This is a very simple tweak, but I imagine with some thought and creativity, some really interesting things might crop up. Also, please contact me if you've played with R for generating other fractal objects, dynamical systems and chaotic systems and the likes. I'd be interested to see what other people have come up with.   


